---
title: "Centering Variables, ANOVA Tables and ANCOVA"
subtitle: "Why and how to center variables when generating ANOVA tables, and examples"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(afex)
library(tidyverse)
library(kableExtra)
afex_options(es_aov = 'pes',
             correction_aov = 'GG',
             emmeans_model = 'univariate')
Alcohol_data <- readRDS("Alcohol_data.rds")
knitr::opts_chunk$set(echo = FALSE)
```


## What Are ANOVA Tables?

ANOVA tables, like regression tables, produce significance tests (and sometimes estimates of effect sizes). Unlike regression tables, where a test is given for each coefficient, in ANOVA tables a test is given by some grouping scheme: by model (of model comparison), of by factor where all coefficients that represent a categorical variable are tested in a joint test. It is the latter table, used usually in analysis of factorial data, that is discussed here.

> Thesis: centering predictors changes the results given by ANOVA tables.

Generally, the results given by ANOVA tables with centered variables are the ones we are interested in.

### Why Center?

In moderation models / models with interaction terms, centering of variables affects the estimates (and thus the joint test and significance) of lower order terms ([Dalal & Zickar, 2011](https://doi.org/10.1177%2F1094428111430540)). It is only after centering variables, that these tests for lower order terms represent what we expect them to - *main effects*, across the levels of all other terms on average ([AFNI](https://afni.nimh.nih.gov/pub/dist/doc/htmldoc/STATISTICS/center.html#centering-with-one-group-of-subjects)).

### How to Center Variables?

*Centering* is mathematical process that gives the 0 of variable $X$ some non-arbitrary meaning. This can be any value, but usually we are interested in the mean of $X$.

### Continuous Variables

Centering of continuous variables is pretty straightforward - we subtract the mean of $X$ from each value of $X$:

$$
X_{centered} = X-\bar{X}
$$
Note that we don't have to subtract the mean; for example, if $X$ is IQ, 0 is meaningless - in fact, it's not even on the scale (with a lower bound of ~50)! I can subtract the mean of my sample, but I can also subtract 100 instead, which is the "population average". Similarly, if $X$ is "age", 0 is a day-old baby, a value that is not usually particularly meaningful.

### Categorical Variables

It would seem line an impossible task - how can you subtract any numeric value from a categorical variable? This is true, but the idea of a "*meaningful 0*" is that in our model giving a value of 0 to this $X$ will represent the average across all level of $X$. When modeling categorical variables, this means setting all dummy variables to 0.

Usually, dummy variables are generated using a treatment scheme, such that 0 represents some "baseline" group. But there are other coding schemes, some of which give 0 the meaning we're looking for - for example, [effects coding](https://stats.idre.ucla.edu/other/mult-pkg/faq/general/faqwhat-is-effect-coding/) ([Aiken, et al., 1991, pp. 127](https://books.google.co.il/books?hl=iw&lr=&id=LcWLUyXcmnkC); [Singmann & Kellen, 2017, pp. 24](http://singmann.org/download/publications/singmann_kellen-introduction-mixed-models.pdf)).^[Unfortunately, this makes the interpretation of the actual coefficient not as straightforward as when using treatment coding...] ^[When conducting ANOVAs with [`afex`](https://cran.r-project.org/package=afex), you don't need to worry about setting effects coding, as `afex` [takes care of this for you](https://github.com/singmann/afex/issues/63). However when conducting ANCOVAs with `afex`, continuous variables [**are not centered**](https://github.com/singmann/afex/issues/59) (but a warning is given), and you have to do that yourself. This is also true for JASP (that has `afex` under the hood) and even for SPSS (but if you're still using SPSS, this might be the least of your problems...).]

In `R` this can be done by setting

```{r, eval=FALSE}
options(contrasts=c('contr.sum', 'contr.poly'))
```

Or for a single factor:

```{r, eval=FALSE}
contrasts(iris$Species) <- contr.sum
```


### Addendum

When generating ANOVA tables, the $SS$ of factor $A$ are computed with all other coefficients held constant at 0 (similar to how coefficients are interpreted as simple slopes in moderation analysis). If factor $B$ has a treatment coding scheme, then when the coefficients of factor $B$ are 0, there actually represent the baseline group, as so the effect for $A$ is actually the simple effect of $A$ and not the main effect!


"But wait!", I hear you shout, "When we learned ANOVA in Intro to stats, we weren't taught to center any variables!".

This is true - you didn't explicitly learn this, but taking a closer look at the equations of the various $SS$s will reveal that you've been doing just that all along. 

For $SS_A$:
$$
SS_A = n\sum (\bar{X}_{i.}-\bar{X}_{..})^2
$$
When $\bar{X}_{i.}$ is itself the mean of group $i$ of factor $A$ *beyond* the levels of factor $B$ (as denoted by the $.$)!

We can even re-write the equation for $SS_A$ to show this explicitly:

$$
SS_A = n\sum (\bar{X}_{i.}-\bar{X}_{..})^2 = 
n\sum\sum (\bar{X}_{ij}-(\bar{X}_{.j}-\bar{X}_{..})-(\bar{X}_{ij}-\bar{X}_{.j}-\bar{X}_{ij}+\bar{X}_{..})-\bar{X}_{..})^2
$$
Where $(\bar{X}_{.j}-\bar{X}_{..})$ is the centering of factor $B$ (subtracting from *all* group means, the total mean).

This is also why centering a variable only affects the lower-order effects of *other* variables.



### Exercise 

*Here's a simple exercise with an empty code chunk provided for entering the answer.*

Write the R code required to add two plus two:

```{r two-plus-two, exercise=TRUE}

```

### Exercise with Code

*Here's an exercise with some prepopulated code as well as `exercise.lines = 5` to provide a bit more initial room to work.*

Now write a function that adds any two numbers and then call it:

```{r add-function, exercise=TRUE, exercise.lines = 5}
add <- function() {
  
}
```

## Fit the ANOVA model

Let's load the data using `Alcohol_data <- readRDS("Alcohol_data.rds")` - already done in the header - and have a look at the data using the code you learned in the last tutorial! 

### Look at the data

Modify the following code to limit the number of rows printed to 6:

```{r data-examine, exercise=TRUE, exercise.eval=TRUE}
Alcohol_data
```

```{r data-examine-hint}
head(Alcohol_data)
```


### Fit the model

OK, let's go ahead and fit the ANOVA model using `aov_ez`. We're going to call the model `ersp_anova`, because the main outcome variable is the ERSP (event-related synaptic potential, an EEG measure). The within-subject factors are Freqency and Correctness, and the between-subjects factor is Alcohol. 


```{r fit-model, exercise = TRUE}
ersp_anova <- aov_ez('Subject', 'ersp', Alcohol_data,
                     within = c(), 
                     between = c())

```


```{r fit-model-solution}
ersp_anova <- aov_ez('Subject','ersp',Alcohol_data,
                     within = c('Frequency','Correctness'),
                     between = c('Alcohol'))
ersp_anova

```

But mother's education level is related to the outcome.

We probably would want to control for it - to reduce the MSE.

### Fit ANCOVA model

Keep in mind that some have argued that the use (or misuse) of ANCOVA should be avoided. See: [here](http://doi.org/10.1037//0021-843X.110.1.40) for more details. 

Copy the code from above, but call the new model `ersp_ancova`, and add in the covariate `mograde`. Be sure to set `factorize = FALSE` at the end! 

```{r fit-ancova-model, exercise = TRUE}


```


```{r fit-ancova-model-solution}
ersp_ancova <- aov_ez('Subject','ersp',Alcohol_data,
                      within = c('Frequency','Correctness'),
                      between = c('Alcohol'),
                      # The new bits:
                      covariate = 'mograde',
                      factorize = FALSE) # MUST set `factorize = FALSE`!
ersp_ancova
```
Note the warning! What does this mean, and how will we fix it? 


### Center the covariate 

Let's make a new column with `mograde` centered - we'll used the `scale` command, and set `center` to `TRUE` but `scale` to `FALSE`: 

```{r center-covariate, exercise = TRUE, exercise.eval = TRUE}
Alcohol_data$mograde_c <- scale(Alcohol_data$mograde,
                                center = TRUE, scale = FALSE)

```


### Re-fit the model

OK, so now let's re-fit the model from above, but with the centered covariate instead! you need to use your new variable `mograde_c`. Amend the code from above, but now add the new centered covariate. 


```{r fit-ancova-model-2, exercise = TRUE, exercise.eval = TRUE}


```


```{r fit-ancova-model-2-solution, exercise.eval = TRUE}
ersp_ancova_2 <- aov_ez('Subject','ersp',Alcohol_data,
                      within = c('Frequency','Correctness'),
                      between = c('Alcohol'),
                      # The new bits:
                      covariate = 'mograde_c',
                      factorize = FALSE) # MUST set `factorize = FALSE`!
ersp_ancova_2
```

Huh! Look at the difference! 

### Make the table prettier

It's nice to present the table in a more readable format. We're going to use the `kableextra` package, along with `tidyverse`, to put the ANOVA table into a data frame and format it nicely as a table. 


```{r kable-styling, exercise = TRUE}

ersp_ancova_table <- as.data.frame(ersp_ancova_2$anova_table) # force ANOVA table into data frame

ersp_ancova_table %>%
  kable(caption = "ANCOVA results with centered covariate") %>%
  kable_styling()
```

### Quiz

*You can include any number of single or multiple choice questions as a quiz. Use the `question` function to define a question and the `quiz` function for grouping multiple questions together.*

Some questions to verify that you understand the purposes of various base and recommended R packages:

```{r quiz}
quiz(
  question("Which package contains functions for installing other R packages?",
    answer("base"),
    answer("tools"),
    answer("utils", correct = TRUE),
    answer("codetools")
  ),
  question("Which of the R packages listed below are used to create plots?",
    answer("lattice", correct = TRUE),
    answer("tools"),
    answer("stats"),
    answer("grid", correct = TRUE)
  )
)
```

