---
output: 
  html_document:
    includes:
      in_header: test1.html
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyr)
library(afex)
library(emmeans)
library(ggplot2)
angle_noise <- read.csv('angle-noise_wide.csv')
data(angle_noise)
knitr::opts_chunk$set(echo = FALSE)
```

# ANOVA made easy: Section 1

## Importing and preparing the data 

For ANOVAs we need to prepare our data in two ways:

 1. If we have many observations per subject / condition, we must aggregate
   the data to a single value per subject / condition. This can be done with,
   for example, one of the following:
   
`aggregate()`
  
`dplyr`'s `summarise()`
  
`prepdat`'s `prep()`
  
(Note: if you're using (G)LMMs you can, but don't have to aggregate.
    
 2. The data must be in the long format.
 
 We have imported the data at the top using the code `angle_noise <- read.csv('angle-noise_wide.csv')` - but it is in wide format

### Wide vs long data.

 WIDE DATA has:
 
 1. A row for each subject,
 
 2. Between-subject variables have a column
 
 3. Repeated measures are stored across columns, and the within-subject are stored in column names


### Exercise 1: Look at the data (with Hint)

Let's have a look at this dataset. If we just type in the name of the data, we can see a whole bunch of rows and columns. Let's look at just the first few rows. 

Modify the following code to limit the number of rows printed to 6:

```{r print-limit, exercise=TRUE, exercise.eval=TRUE}
angle_noise
```

```{r print-limit-hint}
head(angle_noise)
```
 
<div data-iframe-height></div>